Geometric models overestimate lake depth due to imperfect slope measurement

### 2020-11-05 manuscript/agujournaltemplate.tex
more minor text edits

All lake bathymetry was specifically calculated relative to high-resolution (1:24,000 scale) NHD \cite{usgsNationalHydrographyDataset2019} waterbodies such that source data and bathymetry surface outputs were clipped to the area of each lake polygon. We restricted the lakes in our study to those with an area of at least 4 ha and a maximum depth of at least 0.3 m (1 ft). The purpose of these restrictions was to ensure that lakes had enough contours (or points, or polygons) to generate adequately smooth interpolations with which to calculate in-lake geometry metrics.

We used our generated bathymetry surfaces to find the location of the deepest point in the lake and we resolved ties by choosing the deepest point that was closest to the center of the lake. We calculated the center of the lake not as its centroid but rather by finding the point farthest from the lake shoreline (i.e. its “visual distance to lake center”). We did this to account for lakes where the centroid does not intersect lake bathymetry because it is located within an embedded island or peninsula. For these calculations, we used the \texttt{polylabelr} R package \cite{larssonPolylabelrFindPole2019}, which interfaces with the Mapbox pole of inaccessibility algorithm \cite{agafonkinJSLibraryFinding2019}. We calculated in-lake slope as maximum lake depth divided by the distance to the deepest point. Where maximum lake depth is a point measurement derived from a bathymetric surface and distance to the deepest point is the smallest straight-line distance from the waterbody perimeter.

@@

We further classified lakes using the output of a deep convolutional neural network model trained on sattelite images labeled according to whether there was visual evidence of a water control structure significantly impacting flow. This model had an overall validation accuracy of 81\% and produced a probability for each waterbody as to whether it is a reservoir or a natural lake. For our purposes, we set a conservative classification probability threshold of 0.75 to determine whether a lake would be considered a  reservoir. We tested the sensitivity of our model output relative to variations of this threshold and found little substantive difference in results. Note that our reservoir classification defines reservoirs as any permanent waterbody that has a water control structure likely to significantly impact flow or pool water. It makes no distinction between different dam types, heights, or uses/purposes.

@@

\noindent
where sinuous lakes have larger values of shoreline development and circular lakes have smaller values of shoreline development. Watershed to lake area ratio is an approximation of water residence time and is defined as watershed area divided by lake area.

@@

\subsubsection{Random forest models}
\noindent
Prior studies using geometric models to predict lake depth include a statistical or machine learning model “layer” or “offset” to boost predictive accuracy. For our purposes, this offset modeling enabled us to test our expectations that prediction error would be different among different formulations of $depth_{geometric}$ and among different lake types. It also facilitated direct comparison against prior models of lake depth including those that are non-geometric. We generated an “offset” to geometric depth \cite<sensu>{hollisterPredictingMaximumLake2011} using the random forest algorithm and the \texttt{ranger} R package \cite{wrightRangerFastImplementation2017} to predict observed maximum depth as a function of covariates including geometric maximum depth (from Equation \ref{eq1}) along with the lake elevation, area, perimeter, and ratio/index measures listed in Table 1:

@@

\subsection{Representativeness of proxy measures of lake geometry}
\noindent
In comparing among lake geometry measures, our analysis suggests that proxy distance to lake center is representative of true distance to the deepest point of the lakes but that proxy nearshore land slope is not representative of true in-lake slope. A simple indication of this non representativeness is that proxy nearshore land slope was often (in $>$ 74\% of cases) steeper than true in-lake slope. This finding is consistent with \citeA{heathcotePredictingBathymetricFeatures2015} whos results suggest that in-lake slopes are shallower compared to the surrounding land. The shallow nature of in-lake slopes is likley a function of erosion and sediment transport processes \cite{hakansonLakeBottomDynamics1981, johanssonNewApproachesModelling2007}.

@@

\section{Conclusion}
\noindent
To our knowledge, the present study is the largest and most comprehensive test to date of geometric models of lake depth. Using bathymetry data on approximately 5,000 lakes, we show that proxy slope measures are not representative of true in-lake slope and this leads to inaccuracies in predicting the depth of concave and reservoir lakes. These innaccuracies suggest that caution is warranted in using geometric models for depth prediction in unsampled lakes. Despite these apparent biases, overall prediction accuracy was equivalent to that of prior depth prediction studies ($\pm$ 6-7m). Only our models using a true measure of in-lake slope, which is limited in availability to lakes with bathymetry data and where we already know lake depth, had greater accuracy than that of prior studies ($\pm$ 4.2m). Lack of improved prediction accuracy (short of including data that is unavailable for most lakes) suggests that improved prediction may require new types of data or novel analysis techniques.

>>>>

All lake bathymetry was specifically calculated relative to high-resolution (1:24,000 scale) National Hydrography Dataset \cite{usgsNationalHydrographyDataset2019} waterbodies such that source data and bathymetry surface outputs were clipped to the area of each lake polygon. We restricted the lakes in our study to those with an area of at least 4 ha and a maximum depth of at least 0.3 m (1 ft). The purpose of these restrictions was to ensure that lakes had enough contours (or points, or polygons) to generate adequately smooth interpolations with which to calculate in-lake geometry metrics.

We used our generated bathymetry surfaces to find the location of the deepest point in the lake and we resolved ties by choosing the deepest point that was closest to the center of the lake. We calculated the center of the lake not as its centroid but rather by finding the point farthest from the lake shoreline (i.e. its “visual distance to lake center”). We did this to account for lakes where the centroid does not intersect lake bathymetry because it is located within an embedded island or peninsula. For these calculations, we used the \texttt{polylabelr} R package \cite{larssonPolylabelrFindPole2019}, which interfaces with the Mapbox pole of inaccessibility algorithm \cite{agafonkinJSLibraryFinding2019}. We calculated in-lake slope as maximum lake depth divided by the distance to the deepest point where maximum lake depth is a point measurement derived from a bathymetric surface and distance to the deepest point is the smallest straight-line distance from that point to the waterbody perimeter.

@@

We further classified lakes using the output of a deep convolutional neural network model trained on satellite images labeled according to whether there was visual evidence of a water control structure significantly impacting flow. This model had an overall validation accuracy of 81\% and produced a probability for each waterbody as to whether it is a reservoir or a natural lake. For our purposes, we set a conservative classification probability threshold of 0.75 to determine whether a lake would be considered a  reservoir. We tested the sensitivity of our model output relative to variations of this threshold and found little substantive difference in results. Note that our reservoir classification defines reservoirs as any permanent waterbody that has a water control structure likely to significantly impact flow or pool water. It makes no distinction between different dam types, heights, or uses/purposes.

@@

\noindent
where sinuous lakes have larger values of shoreline development and circular lakes have smaller values of shoreline development. Watershed to lake area ratio is an approximation of water residence time and is defined as watershed area divided by lake area \cite{timmsGeomorphologyLakeBasins2009}.

@@

\subsubsection{Random forest models}
\noindent
Prior studies using geometric models to predict lake depth include a statistical or machine learning model “layer” or “offset” to boost predictive accuracy  \cite{hollisterPredictingMaximumLake2011, yigzawNewGlobalStorage2018}. For our purposes, this offset modeling enabled us to test our expectations that prediction error would be different among different formulations of $depth_{geometric}$ and among different lake types. It also facilitated direct comparison against prior models of lake depth including those that are non-geometric. We generated an “offset” to geometric depth \cite<sensu>{hollisterPredictingMaximumLake2011} using the random forest algorithm and the \texttt{ranger} R package \cite{wrightRangerFastImplementation2017} to predict observed maximum depth as a function of covariates including geometric maximum depth (from Equation \ref{eq1}) along with the lake elevation, area, perimeter, and ratio/index measures listed in Table 1:

@@

\subsection{Representativeness of proxy measures of lake geometry}
\noindent
In comparing among lake geometry measures, our analysis suggests that proxy distance to lake center is representative of true distance to the deepest point of the lakes but that proxy nearshore land slope is not representative of true in-lake slope. A simple indication of this non representativeness is that proxy nearshore land slope was often (in $>$ 74\% of cases) steeper than true in-lake slope. This finding is consistent with \citeA{heathcotePredictingBathymetricFeatures2015} whos results suggest that in-lake slopes are shallower compared to the surrounding land. The reason for these shallow in-lake slopes may be related to slope-induced turbidity currents which distribute sediment from shallow high-energy areas of lakes to deep low-energy areas \cite{hakansonLakeBottomDynamics1981, johanssonNewApproachesModelling2007}. The strength of such sediment focusing is likely greater in "younger" lakes with steeper slopes leading to a smoothing of their bathymetry over time \cite{blaisInfluenceLakeMorphometry1995}.

@@

\section{Conclusion}
\noindent
To our knowledge, the present study is the largest and most comprehensive test to date of geometric models of lake depth. Using bathymetry data on approximately 5,000 lakes, we show that proxy slope measures are not representative of true in-lake slope and this leads to inaccuracies in predicting the depth of concave and reservoir lakes. These inaccuracies suggest that caution is warranted in using geometric models for depth prediction in unsampled lakes. Despite these apparent biases, overall prediction accuracy was equivalent to that of prior depth prediction studies ($\pm$ 6-7m). Only our models using a true measure of in-lake slope, which is limited in availability to lakes with bathymetry data and where we already know lake depth, had greater accuracy than that of prior studies ($\pm$ 4.2m). Lack of improved prediction accuracy (short of including data that is unavailable for most lakes) suggests that improved prediction may require new types of data or novel analysis techniques.
